---
title: "Lab 5: Decision Tress"
author: "Juliana Vinluan"
date: "2025-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

There is a discrepancy between "not survived" and "not alive" patients. 
"Not survived" refers to patients who did not survive beyond the study's specified time frame, whereas "not alive" includes all deceased patients, regardless of the time frame.
The distinction is important for modeling survivability rates.

During preprocessing, the authors handled missing data, removed irrelevant features, and converted categorical variables into numerical form. 
They also normalized continuous data and ensured class balance where needed.

The authors used information gain and statistical tests to evaluate the predictive power of different attributes. 
Features with higher information gain were considered more influential in predicting survivability.

Best-performing learning algorithm:

The C4.5 decision tree algorithm outperformed other models, such as Naïve Bayes and backpropagation neural networks, in terms of accuracy and interpretability.

Yes, the dataset contained missing values. The authors handled this by either imputing missing values or excluding incomplete records when necessary.


##### First, I will import the data set and load a library package to use later in this report.


```{r load-packages}
# Import data set
cardata <- read.table("car.data", header = FALSE, sep = ",", stringsAsFactors = TRUE)
healthydata <- read.csv("NPHA-doctor-visits.csv", header = TRUE)

# Include library
library(rpart) # Decision tree algorithm
library(rpart.plot) # Decision tree visualization
library(caret) # Classification and regression training
library(ggplot2)  # Visualization
library(dplyr)    # Data manipulation

# Assign column names
colnames(cardata) <- c("buying", "maint", "doors", "persons", "lug_boot", "safety", "acceptability")
colnames(healthydata) <- c("visits", "age", "phyiscal_health", "mental_health", "dental_health", "employment", "stress_prevents_sleeping", "medication_prevents_sleeping", "pain_prevents_sleeping", "bathroom_prevents_sleeping", "unknown_prevents_sleeping", "trouble_sleeping", "prescription_sleep_meds", "race", "gender")
```

## Data Preprocessing
```{r check-data}
head(cardata)   # View the first few rows
str(cardata)    # Check the structure (columns, types)
dim(cardata)    # Check dimensions (rows, columns)
# Check for missing values
colSums(is.na(cardata))
```
```{r}
# Convert categorical variables to factors
cardata[] <- lapply(cardata, as.factor)

# Split into training and test sets
set.seed(123)
trainIndex <- caret::createDataPartition(cardata$acceptability, p = 0.7, list = FALSE)
trainSet <- cardata[trainIndex, ]
testSet <- cardata[-trainIndex, ]
```

## Build Decision Tree
```{r}
tree_model <- rpart(acceptability ~ ., data = trainSet, method = "class",
                    parms = list(split = "information"))
```

## Visualize Decision Tree
```{r}
rpart.plot(tree_model)
```

## Compute Classification Error *before pruning*
```{r}
pred_train <- factor(predict(tree_model, trainSet, type = "class"), levels = levels(trainSet$acceptability))
pred_test <- factor(predict(tree_model, testSet, type = "class"), levels = levels(trainSet$acceptability))

train_error <- mean(pred_train != trainSet$acceptability)
test_error <- mean(pred_test != testSet$acceptability)

cat("Training Error:", train_error, "\n")
cat("Test Error:", test_error, "\n")
```

## Prune the Tree
```{r}
printcp(tree_model) # Display complexity parameter table
plotcp(tree_model)  # Visualize cp graph
# Choose optimal cp value (e.g., lowest cross-validation error)
best_cp <- tree_model$cptable[which.min(tree_model$cptable[, "xerror"]), "CP"] #After finding the index of the minimum xerror, this line extracts the corresponding CP value.
pruned_tree <- prune(tree_model, cp = best_cp) # Prune the tree using the best CP value
```

## Visualize/Compare Original Tree and Pruned Tree
```{r}
rpart.plot(tree_model)
```
```{r}
rpart.plot(pruned_tree)
```

## Compute Classification Error *after pruning*
```{r}
pred_train_pruned <- factor(predict(pruned_tree, trainSet, type = "class"), levels = levels(trainSet$acceptability))
pred_test_pruned <- factor(predict(pruned_tree, testSet, type = "class"), levels = levels(trainSet$acceptability))

train_error_pruned <- mean(pred_train_pruned != trainSet$acceptability)
test_error_pruned <- mean(pred_test_pruned != testSet$acceptability)

cat("Pruned Training Error:", train_error_pruned, "\n")
cat("Pruned Test Error:", test_error_pruned, "\n")
```

## Repeat with Gini Index
```{r}
tree_model_gini <- rpart(acceptability ~ ., data = trainSet, method = "class", 
                    parms = list(split = "gini"))
```

## Visualize Decision Tree (Gini)
```{r}
rpart.plot(tree_model_gini)
```

## Compute Classification Error (Gini)
```{r}
pred_train_gini <- predict(tree_model_gini, trainSet, type = "class")
pred_test_gini <- predict(tree_model_gini, testSet, type = "class")

train_error_gini <- mean(pred_train_gini != trainSet$acceptability)
test_error_gini <- mean(pred_test_gini != testSet$acceptability)

cat("Training Error (Gini):", train_error_gini, "\n")
cat("Test Error (Gini):", test_error_gini, "\n")
```

## Prune the Tree (Gini)
```{r}
printcp(tree_model_gini) # Display complexity parameter table
plotcp(tree_model_gini)  # Visualize cp graph

# Choose optimal cp value (e.g., lowest cross-validation error)
best_cp_gini <- tree_model_gini$cptable[which.min(tree_model_gini$cptable[, "xerror"]), "CP"]
pruned_tree_gini <- prune(tree_model_gini, cp = best_cp_gini)
print(best_cp_gini)
```

## Visualize Pruned Tree (Gini)
```{r}
rpart.plot(pruned_tree_gini)
```

## Comparison and Conclusion
Summarize the results and discuss whether there are substantial differences between using Information Gain and Gini Index. Note any trends or surprising outcomes.

## Data Preprocessing for Healthy Data
```{r check-healthy-data}
head(healthydata)   # View the first few rows
str(healthydata)    # Check the structure (columns, types)
dim(healthydata)    # Check dimensions (rows, columns)
# Check for missing values
colSums(is.na(healthydata))
```
```{r}
# Convert categorical variables to factors
healthydata[] <- lapply(healthydata, as.factor)

# Split into training and test sets
set.seed(123)
trainIndex2 <- createDataPartition(healthydata$visits, p = 0.7, list = FALSE)
trainSet2 <- healthydata[trainIndex2, ]
testSet2 <- healthydata[-trainIndex2, ]
```


## Build Decision Tree (Information Gain)
```{r}
tree_model_healthy <- rpart(visits ~ ., data = trainSet2, method = "class",
                    parms = list(split = "information"))
ls()
```

## Visualize Decision Tree
```{r}
rpart.plot(tree_model_healthy, box.palette = "Blues")
```

## Compute Classification Error
```{r}
pred_trained_info <- factor(predict(tree_model_healthy, trainSet2, type = "class"), 
                          levels = levels(trainSet2$visits))
pred_testing_info <- factor(predict(tree_model_healthy, testSet2, type = "class"), 
                         levels = levels(testSet2$visits))

trained_error_info <- mean(pred_trained_info != trainSet2$visits)
testing_error_info <- mean(pred_testing_info != testSet2$visits)

cat("Training Error (Info Gain):", trained_error_info, "\n")
cat("Test Error (Info Gain):", testing_error_info, "\n")
```

## Prune the Tree
```{r}
printcp(tree_model_healthy) # Display complexity parameter table
plotcp(tree_model_healthy)  # Visualize cp graph

# Choose optimal cp value (e.g., lowest cross-validation error)
best_cp_info <- tree_model_healthy$cptable[which.min(tree_model_healthy$cptable[, "xerror"]), "CP"]
pruned_tree_healthy <- prune(tree_model_healthy, cp = best_cp_info)

```

## Visualize Pruned Tree
```{r}
rpart.plot(pruned_tree_healthy, box.palette = "Blues")
```

## Compute Classification Error After Pruning
```{r}
pred_train_pruned_info <- predict(pruned_tree_healthy, trainSet2, type = "class")
pred_test_pruned_info <- predict(pruned_tree_healthy, testSet2, type = "class")

train_error_pruned_info <- mean(pred_train_pruned_info != trainSet2$visits)
test_error_pruned_info <- mean(pred_test_pruned_info != testSet2$visits)

cat("Pruned Training Error (Info Gain):", train_error_pruned_info, "\n")
cat("Pruned Test Error (Info Gain):", test_error_pruned_info, "\n")
```

## Conclusion
The pruned decision tree collapsed to just a root node, it may suggest:
The Predictive Power of Features is Weak
The features may not have strong relationships with the target variable.
Check the variable importance to see if any features are contributing to the model.

```{r}
varImp(tree_model_healthy)
```

What This Tells You About the Data
Employment, Medication, and Physical Health are key predictors of sleep issues.

Demographics (Race) and Prescription Sleep Meds also contribute significantly.

Age and Gender had no impact.

Possible Next steps:
Feature Selection: I might consider dropping age and gender since they had no effect.

Further Analysis: Investigate why pain_prevents_sleeping had such a low score—does the dataset properly capture pain-related sleep disturbances?

Try a Different Model: Another model like Random Forest might yield better results.